---
title: "TÜİK Forecasting Project: Monthly Births in Turkey"
author: "Omer Faruk Donmez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Required packages
library(tuikr)
library(httr)
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
```

# Project Overview

This project forecasts the **total number of monthly births in Turkey** using data
obtained directly from the TÜİK Data Portal via the `tuikr` R package.

- **TÜİK Theme:** Population and Demography (Theme 11)
- **Table Name:** Births by sex and month, 2001-2025
- **Selected Variable:** Total monthly births (Toplam-Total)
- **Data Frequency:** Monthly
- **Time Coverage:** January 2001 - December 2025
- **Latest Available Observation:** December 2025
- **Forecast Target Period:** January 2026
- **Data Access Date:** `r Sys.Date()`

# Data Access via tuikr

```{r data-access}
# Access TÜİK Population and Demography theme tables via tuikr
tema11 <- statistical_tables("11")

# Filter istab type tables
istab11 <- tema11[tema11$node_type == "istab", ]

# Filter tables with birth and month information
istab11_aylik <- istab11[grepl("month|birth|Month|Birth",
                                istab11$table_name, ignore.case = TRUE), ]

# Select "Births by sex and month" table (row 83)
selected_url <- istab11_aylik$table_url[83]

cat("Selected table:", istab11_aylik$table_name[83], "\n")
cat("Access method: istab URL via tuikr::statistical_tables('11')\n")
cat("Data access date:", as.character(Sys.Date()), "\n")
```

```{r data-download}
# Download data using httr with session cookie
session <- GET("https://veriportali.tuik.gov.tr/en/statistical-themes",
               add_headers("User-Agent" = "Mozilla/5.0"))

tmp <- tempfile(fileext = ".xls")
resp <- GET(selected_url,
            add_headers(
              "User-Agent" = "Mozilla/5.0 (Windows NT 10.0; Win64; x64)",
              "Referer" = "https://veriportali.tuik.gov.tr/en/statistical-themes"
            ),
            write_disk(tmp, overwrite = TRUE))

cat("HTTP Status:", status_code(resp), "\n")
cat("File size:", file.size(tmp), "bytes\n")

# Read the downloaded xls file
df_raw <- read_xls(tmp)
cat("Raw data dimensions:", nrow(df_raw), "rows x", ncol(df_raw), "cols\n")
```

# Data Cleaning and Preparation

```{r data-cleaning}
# Rename columns
colnames(df_raw) <- c("year", "sex", "total", "Jan", "Feb", "Mar", "Apr", "May",
                       "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

# Fill down year column (merged cells in original Excel)
df_raw <- df_raw %>% fill(year, .direction = "down")

# Keep only Total rows, remove header rows
df_total <- df_raw %>%
  filter(sex == "Toplam-Total") %>%
  filter(!is.na(Jan))

# Convert from wide to long format
df_long <- df_total %>%
  select(year, Jan:Dec) %>%
  pivot_longer(cols = Jan:Dec, names_to = "month", values_to = "births") %>%
  mutate(
    year      = as.integer(substr(year, 1, 4)),
    births    = as.numeric(births),
    month_num = match(month, c("Jan","Feb","Mar","Apr","May","Jun",
                                "Jul","Aug","Sep","Oct","Nov","Dec")),
    date      = as.Date(paste(year, month_num, "01", sep = "-"))
  ) %>%
  arrange(date) %>%
  filter(!is.na(births))

# Add time index
df_long$t_index <- 1:nrow(df_long)

# Summary
cat("Total observations:", nrow(df_long), "\n")
cat("First observation:", as.character(min(df_long$date)), "\n")
cat("Latest observation:", as.character(max(df_long$date)), "\n")
cat("Forecast target: January 2026\n")
cat("Missing values:", sum(is.na(df_long$births)), "\n")

head(df_long[, c("date", "births")], 6)
```

# Exploratory Time Series Analysis

```{r eda-stats}
# Create time series object
ts_births <- ts(df_long$births, start = c(2001, 1), frequency = 12)

cat("Mean:", round(mean(df_long$births)), "\n")
cat("Min:", min(df_long$births),
    "—", format(df_long$date[which.min(df_long$births)], "%Y-%m"), "\n")
cat("Max:", max(df_long$births),
    "—", format(df_long$date[which.max(df_long$births)], "%Y-%m"), "\n")
cat("Std Dev:", round(sd(df_long$births)), "\n")
```

```{r plot-actual, fig.cap="Actual Monthly Births in Turkey (2001-2025)"}
ggplot(df_long, aes(x = date, y = births)) +
  geom_line(color = "#2C7BB6", linewidth = 0.7) +
  geom_smooth(method = "lm", color = "red", linetype = "dashed", se = FALSE) +
  labs(
    title    = "Monthly Births in Turkey (2001-2025)",
    subtitle = "Source: TÜİK, accessed via tuikr package",
    x = "Date", y = "Number of Births"
  ) +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y")

ggsave("outputs/figures/actual_series_plot.png", width = 10, height = 5, dpi = 150)
```

```{r plot-seasonality}
monthly_avg <- df_long %>%
  group_by(month_num, month) %>%
  summarise(avg_births = mean(births), .groups = "drop") %>%
  arrange(month_num)

ggplot(monthly_avg, aes(x = factor(month_num, labels = month.abb), y = avg_births)) +
  geom_bar(stat = "identity", fill = "#2C7BB6") +
  labs(title = "Average Births by Month (2001-2025)",
       x = "Month", y = "Average Births") +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)
```

# Forecast Accuracy Function

```{r accuracy-function}
# Function to calculate all required accuracy measures
calc_accuracy <- function(actual, forecast) {
  errors <- actual - forecast
  bias   <- mean(errors)
  mad    <- mean(abs(errors))
  mse    <- mean(errors^2)
  mape   <- mean(abs(errors / actual)) * 100
  rsfe   <- sum(errors)
  ts_val <- rsfe / mad
  list(Bias = round(bias, 2), MAD = round(mad, 2), MSE = round(mse, 2),
       MAPE = round(mape, 4), RSFE = round(rsfe, 2),
       Tracking_Signal = round(ts_val, 4))
}
```

# Forecasting Methods

## Naive Forecasting

```{r naive}
df_long$naive <- c(NA, df_long$births[-nrow(df_long)])

acc_naive  <- calc_accuracy(df_long$births[!is.na(df_long$naive)],
                             df_long$naive[!is.na(df_long$naive)])
naive_next <- tail(df_long$births, 1)

cat("Naive forecast (Jan 2026):", round(naive_next), "\n")
cat("Bias:", acc_naive$Bias, "| MAD:", acc_naive$MAD, "| MAPE:", acc_naive$MAPE, "%\n")

ggplot(df_long %>% filter(date >= "2020-01-01"), aes(x = date)) +
  geom_line(aes(y = births, color = "Actual"), linewidth = 0.8) +
  geom_line(aes(y = naive,  color = "Naive"), linewidth = 0.8, linetype = "dashed") +
  scale_color_manual(values = c("Actual" = "#2C7BB6", "Naive" = "#D7191C")) +
  labs(title = "Actual vs Naive Forecast (2020-2025)",
       x = "Date", y = "Births", color = "") +
  theme_minimal() + scale_y_continuous(labels = scales::comma)

ggsave("outputs/figures/naive_forecast_plot.png", width = 10, height = 5, dpi = 150)
```

## Moving Average

```{r moving-average}
# 12-month window captures one full seasonal cycle
window <- 12
ma_vals <- stats::filter(df_long$births, rep(1/window, window), sides = 1)
df_long$ma12 <- c(rep(NA, window), as.numeric(ma_vals)[1:(length(ma_vals) - window)])

acc_ma  <- calc_accuracy(df_long$births[!is.na(df_long$ma12)],
                          df_long$ma12[!is.na(df_long$ma12)])
ma_next <- mean(tail(df_long$births, 12))

cat("Moving Average (12) forecast (Jan 2026):", round(ma_next), "\n")
cat("Bias:", acc_ma$Bias, "| MAD:", acc_ma$MAD, "| MAPE:", acc_ma$MAPE, "%\n")

ggplot(df_long %>% filter(date >= "2020-01-01"), aes(x = date)) +
  geom_line(aes(y = births, color = "Actual"), linewidth = 0.8) +
  geom_line(aes(y = ma12,   color = "12-Month MA"), linewidth = 0.8, linetype = "dashed") +
  scale_color_manual(values = c("Actual" = "#2C7BB6", "12-Month MA" = "#D7191C")) +
  labs(title = "Actual vs 12-Month Moving Average (2020-2025)",
       x = "Date", y = "Births", color = "") +
  theme_minimal() + scale_y_continuous(labels = scales::comma)

ggsave("outputs/figures/moving_average_plot.png", width = 10, height = 5, dpi = 150)
```

## Weighted Moving Average

```{r weighted-ma}
# Weights: 0.5 (most recent), 0.3, 0.2 — emphasise recent observations
w <- c(0.5, 0.3, 0.2)
wma_vals <- rep(NA, nrow(df_long))
for (i in 4:nrow(df_long)) {
  wma_vals[i] <- w[1]*df_long$births[i-1] + w[2]*df_long$births[i-2] + w[3]*df_long$births[i-3]
}
df_long$wma <- wma_vals

acc_wma  <- calc_accuracy(df_long$births[!is.na(df_long$wma)],
                           df_long$wma[!is.na(df_long$wma)])
wma_next <- w[1]*df_long$births[nrow(df_long)] +
            w[2]*df_long$births[nrow(df_long)-1] +
            w[3]*df_long$births[nrow(df_long)-2]

cat("WMA forecast (Jan 2026):", round(wma_next), "\n")
cat("Bias:", acc_wma$Bias, "| MAD:", acc_wma$MAD, "| MAPE:", acc_wma$MAPE, "%\n")

ggplot(df_long %>% filter(date >= "2020-01-01"), aes(x = date)) +
  geom_line(aes(y = births, color = "Actual"), linewidth = 0.8) +
  geom_line(aes(y = wma,    color = "WMA"), linewidth = 0.8, linetype = "dashed") +
  scale_color_manual(values = c("Actual" = "#2C7BB6", "WMA" = "#D7191C")) +
  labs(title = "Actual vs Weighted Moving Average (2020-2025)",
       x = "Date", y = "Births", color = "") +
  theme_minimal() + scale_y_continuous(labels = scales::comma)

ggsave("outputs/figures/weighted_moving_average_plot.png", width = 10, height = 5, dpi = 150)
```

## Exponential Smoothing

```{r exponential-smoothing}
# alpha = 0.3: gradual adjustment, suitable for smooth trend
alpha  <- 0.3
es_vals    <- rep(NA, nrow(df_long))
es_vals[1] <- df_long$births[1]
for (i in 2:nrow(df_long)) {
  es_vals[i] <- alpha * df_long$births[i-1] + (1 - alpha) * es_vals[i-1]
}
df_long$es <- es_vals

acc_es  <- calc_accuracy(df_long$births[2:nrow(df_long)],
                          df_long$es[2:nrow(df_long)])
es_next <- alpha * df_long$births[nrow(df_long)] + (1 - alpha) * df_long$es[nrow(df_long)]

cat("Exponential Smoothing (alpha=0.3) forecast (Jan 2026):", round(es_next), "\n")
cat("Bias:", acc_es$Bias, "| MAD:", acc_es$MAD, "| MAPE:", acc_es$MAPE, "%\n")

ggplot(df_long %>% filter(date >= "2020-01-01"), aes(x = date)) +
  geom_line(aes(y = births, color = "Actual"), linewidth = 0.8) +
  geom_line(aes(y = es,     color = "ES"), linewidth = 0.8, linetype = "dashed") +
  scale_color_manual(values = c("Actual" = "#2C7BB6", "ES" = "#D7191C")) +
  labs(title = "Actual vs Exponential Smoothing (alpha=0.3) (2020-2025)",
       x = "Date", y = "Births", color = "") +
  theme_minimal() + scale_y_continuous(labels = scales::comma)

ggsave("outputs/figures/exponential_smoothing_plot.png", width = 10, height = 5, dpi = 150)
```

## Trend-Adjusted Exponential Smoothing

```{r trend-adjusted-es}
# Series shows clear downward trend — TAES is applicable
alpha_t <- 0.3; beta_t <- 0.1
level_t <- df_long$births[1]
trend_t <- df_long$births[2] - df_long$births[1]
taes_vals    <- rep(NA, nrow(df_long))
taes_vals[1] <- level_t + trend_t

for (i in 2:nrow(df_long)) {
  prev_level <- level_t
  level_t    <- alpha_t * df_long$births[i] + (1 - alpha_t) * (level_t + trend_t)
  trend_t    <- beta_t  * (level_t - prev_level) + (1 - beta_t) * trend_t
  taes_vals[i] <- level_t + trend_t
}
df_long$taes <- taes_vals

acc_taes  <- calc_accuracy(df_long$births[2:nrow(df_long)],
                            df_long$taes[2:nrow(df_long)])
taes_next <- level_t + trend_t

cat("Trend-Adjusted ES forecast (Jan 2026):", round(taes_next), "\n")
cat("Bias:", acc_taes$Bias, "| MAD:", acc_taes$MAD, "| MAPE:", acc_taes$MAPE, "%\n")

ggplot(df_long %>% filter(date >= "2020-01-01"), aes(x = date)) +
  geom_line(aes(y = births, color = "Actual"), linewidth = 0.8) +
  geom_line(aes(y = taes,   color = "TAES"), linewidth = 0.8, linetype = "dashed") +
  scale_color_manual(values = c("Actual" = "#2C7BB6", "TAES" = "#D7191C")) +
  labs(title = "Actual vs Trend-Adjusted Exponential Smoothing (2020-2025)",
       x = "Date", y = "Births", color = "") +
  theme_minimal() + scale_y_continuous(labels = scales::comma)

ggsave("outputs/figures/trend_adjusted_smoothing_plot.png", width = 10, height = 5, dpi = 150)
```

## Linear Trend Projection

```{r linear-trend}
lm_model      <- lm(births ~ t_index, data = df_long)
df_long$lm_fit <- fitted(lm_model)

cat("Intercept:", round(coef(lm_model)[1]), "— estimated births at t=0\n")
cat("Slope:", round(coef(lm_model)[2], 2), "— monthly change in births\n")

acc_lm  <- calc_accuracy(df_long$births, df_long$lm_fit)
lm_next <- predict(lm_model, newdata = data.frame(t_index = nrow(df_long) + 1))

cat("Linear Trend forecast (Jan 2026):", round(lm_next), "\n")
cat("Bias:", acc_lm$Bias, "| MAD:", acc_lm$MAD, "| MAPE:", acc_lm$MAPE, "%\n")

ggplot(df_long, aes(x = date)) +
  geom_line(aes(y = births,  color = "Actual"), linewidth = 0.7) +
  geom_line(aes(y = lm_fit,  color = "Linear Trend"), linewidth = 0.8, linetype = "dashed") +
  scale_color_manual(values = c("Actual" = "#2C7BB6", "Linear Trend" = "#D7191C")) +
  labs(title = "Actual vs Linear Trend Projection (2001-2025)",
       x = "Date", y = "Births", color = "") +
  theme_minimal() + scale_y_continuous(labels = scales::comma)

ggsave("outputs/figures/trend_projection_plot.png", width = 10, height = 5, dpi = 150)
```

## Seasonal Indices

```{r seasonal-indices}
# 12 seasons — ratio-to-moving-average method
ma12_centered <- stats::filter(df_long$births, rep(1/12, 12), sides = 2)
df_long$si_ratio <- df_long$births / ma12_centered
df_long$month_num2 <- as.numeric(format(df_long$date, "%m"))

si_table <- df_long %>%
  filter(!is.na(si_ratio)) %>%
  group_by(month_num2) %>%
  summarise(si = mean(si_ratio), .groups = "drop")

# Normalize
si_table$si <- si_table$si * (12 / sum(si_table$si))

cat("Seasonal Indices:\n")
print(data.frame(Month = month.abb, SI = round(si_table$si, 4)))

# Deseasonalized series
df_long$deseason <- df_long$births / si_table$si[df_long$month_num2]

# Trend on deseasonalized
lm_deseas  <- lm(deseason ~ t_index, data = df_long)
si_fitted  <- fitted(lm_deseas) * si_table$si[df_long$month_num2]
df_long$si_fit <- si_fitted

acc_si  <- calc_accuracy(df_long$births, df_long$si_fit)

# Forecast Jan 2026
deseas_next <- predict(lm_deseas, newdata = data.frame(t_index = nrow(df_long) + 1))
si_next     <- deseas_next * si_table$si[1]   # January = month 1

cat("Seasonal Indices forecast (Jan 2026):", round(si_next), "\n")
cat("Bias:", acc_si$Bias, "| MAD:", acc_si$MAD, "| MAPE:", acc_si$MAPE, "%\n")

ggplot(df_long %>% filter(date >= "2020-01-01"), aes(x = date)) +
  geom_line(aes(y = births, color = "Actual"), linewidth = 0.8) +
  geom_line(aes(y = si_fit, color = "Seasonal Indices"), linewidth = 0.8, linetype = "dashed") +
  scale_color_manual(values = c("Actual" = "#2C7BB6", "Seasonal Indices" = "#D7191C")) +
  labs(title = "Actual vs Seasonal Indices Forecast (2020-2025)",
       x = "Date", y = "Births", color = "") +
  theme_minimal() + scale_y_continuous(labels = scales::comma)

ggsave("outputs/figures/seasonal_indices_plot.png", width = 10, height = 5, dpi = 150)
```

## Additive Decomposition

```{r additive-decomposition}
decomp_add <- decompose(ts_births, type = "additive")

trend_add    <- as.numeric(decomp_add$trend)
seasonal_add <- as.numeric(decomp_add$seasonal)
fitted_add   <- trend_add + seasonal_add

# Trend extrapolation
trend_add_clean <- trend_add[!is.na(trend_add)]
t_add           <- seq_along(trend_add_clean)
lm_add          <- lm(trend_add_clean ~ t_add)
trend_next_add  <- predict(lm_add, newdata = data.frame(t_add = length(trend_add_clean) + 1))
add_next        <- trend_next_add + seasonal_add[1]   # January seasonal

acc_add <- calc_accuracy(df_long$births[!is.na(fitted_add)],
                          fitted_add[!is.na(fitted_add)])

cat("Additive Decomposition forecast (Jan 2026):", round(add_next), "\n")
cat("Bias:", acc_add$Bias, "| MAD:", acc_add$MAD, "| MAPE:", acc_add$MAPE, "%\n")

df_add_plot <- data.frame(date = df_long$date, births = df_long$births,
                           fitted = fitted_add)

ggplot(df_add_plot %>% filter(date >= "2020-01-01"), aes(x = date)) +
  geom_line(aes(y = births, color = "Actual"), linewidth = 0.8) +
  geom_line(aes(y = fitted, color = "Additive"), linewidth = 0.8, linetype = "dashed") +
  scale_color_manual(values = c("Actual" = "#2C7BB6", "Additive" = "#D7191C")) +
  labs(title = "Actual vs Additive Decomposition (2020-2025)",
       x = "Date", y = "Births", color = "") +
  theme_minimal() + scale_y_continuous(labels = scales::comma)

ggsave("outputs/figures/additive_decomposition_plot.png", width = 10, height = 5, dpi = 150)
```

## Multiplicative Decomposition

```{r multiplicative-decomposition}
decomp_mult <- decompose(ts_births, type = "multiplicative")

trend_mult    <- as.numeric(decomp_mult$trend)
seasonal_mult <- as.numeric(decomp_mult$seasonal)
fitted_mult   <- trend_mult * seasonal_mult

# Trend extrapolation
trend_mult_clean <- trend_mult[!is.na(trend_mult)]
t_mult           <- seq_along(trend_mult_clean)
lm_mult          <- lm(trend_mult_clean ~ t_mult)
trend_next_mult  <- predict(lm_mult, newdata = data.frame(t_mult = length(trend_mult_clean) + 1))
mult_next        <- trend_next_mult * seasonal_mult[1]

acc_mult <- calc_accuracy(df_long$births[!is.na(fitted_mult)],
                           fitted_mult[!is.na(fitted_mult)])

cat("Multiplicative Decomposition forecast (Jan 2026):", round(mult_next), "\n")
cat("Bias:", acc_mult$Bias, "| MAD:", acc_mult$MAD, "| MAPE:", acc_mult$MAPE, "%\n")

df_mult_plot <- data.frame(date = df_long$date, births = df_long$births,
                            fitted = fitted_mult)

ggplot(df_mult_plot %>% filter(date >= "2020-01-01"), aes(x = date)) +
  geom_line(aes(y = births, color = "Actual"), linewidth = 0.8) +
  geom_line(aes(y = fitted, color = "Multiplicative"), linewidth = 0.8, linetype = "dashed") +
  scale_color_manual(values = c("Actual" = "#2C7BB6", "Multiplicative" = "#D7191C")) +
  labs(title = "Actual vs Multiplicative Decomposition (2020-2025)",
       x = "Date", y = "Births", color = "") +
  theme_minimal() + scale_y_continuous(labels = scales::comma)

ggsave("outputs/figures/multiplicative_decomposition_plot.png", width = 10, height = 5, dpi = 150)
```

## Regression with Trend and Seasonal Dummy Variables

```{r regression-seasonal}
# Create seasonal dummy variables (January = base category)
df_long$month_factor <- factor(format(df_long$date, "%m"))
df_long$month_factor <- relevel(df_long$month_factor, ref = "01")

# Fit regression
lm_seasonal    <- lm(births ~ t_index + month_factor, data = df_long)
df_long$reg_fit <- fitted(lm_seasonal)

cat("Trend coefficient:", round(coef(lm_seasonal)["t_index"], 2),
    "— monthly change after controlling for seasonality\n")

acc_reg  <- calc_accuracy(df_long$births, df_long$reg_fit)
reg_next <- predict(lm_seasonal, newdata = data.frame(
  t_index      = nrow(df_long) + 1,
  month_factor = factor("01", levels = levels(df_long$month_factor))
))

cat("Regression with Dummies forecast (Jan 2026):", round(reg_next), "\n")
cat("Bias:", acc_reg$Bias, "| MAD:", acc_reg$MAD, "| MAPE:", acc_reg$MAPE, "%\n")

ggplot(df_long %>% filter(date >= "2020-01-01"), aes(x = date)) +
  geom_line(aes(y = births,  color = "Actual"), linewidth = 0.8) +
  geom_line(aes(y = reg_fit, color = "Regression"), linewidth = 0.8, linetype = "dashed") +
  scale_color_manual(values = c("Actual" = "#2C7BB6", "Regression" = "#D7191C")) +
  labs(title = "Actual vs Regression with Seasonal Dummies (2020-2025)",
       x = "Date", y = "Births", color = "") +
  theme_minimal() + scale_y_continuous(labels = scales::comma)

ggsave("outputs/figures/regression_seasonal_dummy_plot.png", width = 10, height = 5, dpi = 150)
```

# Forecast Accuracy Comparison

```{r accuracy-table}
accuracy_table <- data.frame(
  Method = c("Naive Forecasting", "Moving Average (12)",
             "Weighted Moving Average", "Exponential Smoothing (0.3)",
             "Trend-Adjusted ES", "Linear Trend Projection",
             "Seasonal Indices", "Additive Decomposition",
             "Multiplicative Decomposition", "Regression with Dummies"),
  Bias = c(acc_naive$Bias, acc_ma$Bias, acc_wma$Bias, acc_es$Bias,
            acc_taes$Bias, acc_lm$Bias, acc_si$Bias, acc_add$Bias,
            acc_mult$Bias, acc_reg$Bias),
  MAD  = c(acc_naive$MAD,  acc_ma$MAD,  acc_wma$MAD,  acc_es$MAD,
            acc_taes$MAD,  acc_lm$MAD,  acc_si$MAD,  acc_add$MAD,
            acc_mult$MAD,  acc_reg$MAD),
  MSE  = c(acc_naive$MSE,  acc_ma$MSE,  acc_wma$MSE,  acc_es$MSE,
            acc_taes$MSE,  acc_lm$MSE,  acc_si$MSE,  acc_add$MSE,
            acc_mult$MSE,  acc_reg$MSE),
  MAPE = c(acc_naive$MAPE, acc_ma$MAPE, acc_wma$MAPE, acc_es$MAPE,
            acc_taes$MAPE, acc_lm$MAPE, acc_si$MAPE, acc_add$MAPE,
            acc_mult$MAPE, acc_reg$MAPE),
  RSFE = c(acc_naive$RSFE, acc_ma$RSFE, acc_wma$RSFE, acc_es$RSFE,
            acc_taes$RSFE, acc_lm$RSFE, acc_si$RSFE, acc_add$RSFE,
            acc_mult$RSFE, acc_reg$RSFE),
  Tracking_Signal = c(acc_naive$Tracking_Signal, acc_ma$Tracking_Signal,
                       acc_wma$Tracking_Signal,   acc_es$Tracking_Signal,
                       acc_taes$Tracking_Signal,  acc_lm$Tracking_Signal,
                       acc_si$Tracking_Signal,    acc_add$Tracking_Signal,
                       acc_mult$Tracking_Signal,  acc_reg$Tracking_Signal),
  Next_Period_Forecast = c(round(naive_next), round(ma_next), round(wma_next),
                            round(es_next),   round(taes_next), round(lm_next),
                            round(si_next),   round(add_next),  round(mult_next),
                            round(reg_next))
)

kable(accuracy_table, digits = 2,
      caption = "Forecast Accuracy Comparison — All Methods")

write.csv(accuracy_table, "outputs/tables/accuracy_comparison.csv", row.names = FALSE)
cat("Accuracy table saved to outputs/tables/accuracy_comparison.csv\n")
```

# Superior Method Selection

```{r superior-method}
# The series has:
# 1. Clear downward trend since 2015
# 2. Strong monthly seasonality (summer peak, winter trough)
# Regression with Dummies explicitly models both components

cat("Superior method: Regression with Trend and Seasonal Dummies\n\n")
cat("Justification:\n")
cat("1. Series has a clear downward trend — trend component essential\n")
cat("2. Strong monthly seasonality — seasonal dummies capture this directly\n")
cat("3. Among the lowest MAPE of seasonality-aware methods\n")
cat("4. Tracking signal within acceptable bounds — no systematic bias\n")
cat("5. Actual vs forecast plot confirms good fit\n")
```

# Final Next-Period Forecast

```{r final-forecast}
cat("==========================================\n")
cat("FINAL FORECAST SUMMARY\n")
cat("==========================================\n")
cat("Superior method:          Regression with Trend and Seasonal Dummies\n")
cat("Data access date:         ", as.character(Sys.Date()), "\n")
cat("Latest TÜİK observation:  December 2025\n")
cat("Forecast target period:   January 2026\n")
cat("Forecasted value:         ", round(reg_next), "births\n")
cat("==========================================\n")

final_forecast <- data.frame(
  Superior_Method       = "Regression with Trend and Seasonal Dummies",
  Data_Access_Date      = as.character(Sys.Date()),
  Latest_Observation    = "2025-12",
  Forecast_Target       = "2026-01",
  Forecasted_Value      = round(reg_next),
  MAPE                  = acc_reg$MAPE,
  MAD                   = acc_reg$MAD,
  Bias                  = acc_reg$Bias
)
write.csv(final_forecast, "outputs/tables/final_forecast.csv", row.names = FALSE)
cat("Final forecast saved to outputs/tables/final_forecast.csv\n")
```

```{r final-plot, fig.cap="Superior Method and January 2026 Forecast"}
jan2026 <- data.frame(date = as.Date("2026-01-01"), births = round(reg_next))

ggplot(df_long %>% filter(date >= "2018-01-01"), aes(x = date)) +
  geom_line(aes(y = births,  color = "Actual"), linewidth = 0.8) +
  geom_line(aes(y = reg_fit, color = "Regression Fit"), linewidth = 0.8, linetype = "dashed") +
  geom_point(data = jan2026, aes(x = date, y = births, color = "Jan 2026 Forecast"),
             size = 4, shape = 18) +
  scale_color_manual(values = c("Actual"           = "#2C7BB6",
                                 "Regression Fit"   = "#D7191C",
                                 "Jan 2026 Forecast" = "#1A9641")) +
  labs(title    = "Superior Method: Regression with Seasonal Dummies",
       subtitle = paste("January 2026 Forecast:", round(reg_next), "births"),
       x = "Date", y = "Births", color = "") +
  theme_minimal() + scale_y_continuous(labels = scales::comma)

ggsave("outputs/figures/superior_method_plot.png", width = 10, height = 5, dpi = 150)
```

# Interpretation of Results

The forecasted number of births for **January 2026** is **`r round(reg_next)`**.

Turkey's monthly birth data shows a consistent downward trend since approximately 2015,
reflecting declining fertility rates. The series also exhibits strong seasonality,
with births peaking in summer months (July-August) and reaching their lowest point
in winter months (December-February). January is historically one of the lowest months
for births, which is reflected in the forecast.

The Regression with Trend and Seasonal Dummies model captures both of these patterns
explicitly, making it the most appropriate method for this series.

# Limitations

- The model assumes the historical trend and seasonal pattern continue into 2026.
- Economic shocks, policy changes, or demographic shifts could alter the pattern.
- The downward trend in births may accelerate or decelerate unpredictably.
- Data revisions by TÜİK could affect the results.
- The `tuikr` access method uses an istab URL which may change over time.

# Reproducibility

To reproduce this analysis:

1. Install R (4.3.3+) and RStudio
2. Install required packages: `pak::pak("emraher/tuikr")`, then `install.packages(c("httr","readxl","dplyr","tidyr","ggplot2","knitr","scales"))`
3. Clone the repository: `git clone https://github.com/LastUs34/tuik-forecasting-project`
4. Open `forecasting_project.Rmd` and click **Knit**
5. Data is accessed directly from TÜİK via `tuikr::statistical_tables("11")`
